Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-05-28 manuscript.qmd
use pip-tools to specify environment
add parapgraphs on data consensus

In this section, we

@@

- all models initialized with the same settings
  - same graph
  - same data
  - same distributions over diagnose time
- parameter samples were drawn for all three models:
  - the sampling used {{< var sampling.walkers_per_dim >}} walkers per parameter space dimension.
  - Every {{< var sampling.check_interval >}} steps, convergence was checked
  - convergence was reached when two criteria were met:
      1. the relative difference between subsequent autocorrelation estimates of the chain does not change more than {{< var sampling.rel_thresh |  >}}.
      2. the autocorrelation estimate of the chain drops below the length of the chain divided by {{< var sampling.trust_fac >}}.
  - after convergence, {{< var sampling.nsteps >}} samples (from the parallel walkers) were drawn that were spaced {{< var sampling.thin >}} steps apart, resulting in `num params x {{< var sampling.walkers_per_dim >}} x {{< var sampling.nsteps >}}` parameter samples.

For every model, we compare the predicted prevalence of a certain scenario to how often this scenario is seen in the data. The considered scenarios are choosen to test if the models can take into account T-category, the correlation of ipsi- and contralateral involvement, and the primary tumor's lateralization (whether or not it extends over the mid-sagittal line).

- [ ] list and describe the scenarios

>>>>

<!-- Maybe I should break this apart into separate chapters on data, sampling, etc. -->

In this section, we detail how the experiments were performed. Every figure, table, and result is fully reproducible via the GitHub repository [`rmnldwg/bilateral-paper`]. It also contains the raw manuscript and instructions on how to recreate all figures, tables, and the final document.

[`rmnldwg/bilateral-paper`]: https://github.com/rmnldwg/bilateral-paper

## Involvement Data Consensus

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference. We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL. When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007].

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

For each dimension in the parameter space of the model, we initialized {{< var sampling.walkers_per_dim >}} of these parallel samplers, called "walkers", with random values in the unit cube. Every time all of these walkers advanced {{< var sampling.check_interval >}} steps, the autocorrelation time of the chains was estimated. For short chains, this estimate is not trustworthy, but stabilizes for longer chains. We therefore considered a sampling to be converged when two criteria where met:

1. The change in the autocorrelation time was less then {{< var sampling.rel_thresh >}}.
2. The estimate of the autocorrelation dropped below $n$ / {{< var sampling.trust_fac >}} where $n$ is the length of the chain up to that point.

All samples up to this convergence - called the *burn-in phase* - were discarded. We only kept another {{< var sampling.nsteps >}} samples after that, which were spaced {{< var sampling.thin >}} steps apart.

[`emcee`]: https://emcee.readthedocs.io/en/stable/
[`lymph-model`]: https://lymph-model.readthedocs.io/en/stable/
